---
title: "Crab-coral model"
author: "Julianna Renzi"
date: "2026-04-08"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Model structure: coral-crabs

First, I'll start with the model we wrote down at WACKME 2026 for coral allocation strategies with crabs:

**Change in host coral (H) over time** = coral growth - coral mortality - predation likelihood * crab defense
$$\frac{dH}{dt} = rC - \ell H - pH\left(\frac{1}{1+D}\right)$$

**Change in host carbon (C) over time** = endosymbiont carbon + heterotrophic carbon - carbon lost to growth - carbon lost to maintenance 
$$\frac{dC}{dt} = eH + hH - rC - mH - fC$$

**Change in crab defensiveness (D) over time** = increase from host investment - weakening
$$\frac{dD}{dt} = afC\left(\frac{sH - D}{sH}\right) - \omega D$$
# Define terms

| Symbol | Type | Description | Units |
|--------|------|-------------|-------|
| $H$ | State variable | Coral biomass | biomass |
| $C$ | State variable | Coral carbon | carbon units |
| $D$ | State variable | Crab defensiveness | defensiveness units |
| $r$ | Parameter | Carbon-to-coral conversion rate (coral growth from carbon) | time$^{-1}$ |
| $\ell$ | Parameter | Coral natural mortality rate | time$^{-1}$ |
| $p$ | Parameter | Probability of predation on coral | time$^{-1}$ |
| $e$ | Parameter | Endosymbiont carbon translocation rate per unit coral  | time$^{-1}$ |
| $h$ | Parameter | Heterotrophic carbon acquisition rate per unit coral (feeding) | time$^{-1}$ |
| $m$ | Parameter | Maintenance carbon cost per coral | time$^{-1}$ |
| $f$ | Parameter | Crab food consumption rate (carbon consumed per unit $C$) | time$^{-1}$ |
| $f$ | Parameter | Investment rate of carbon by coral to crabs in the form of food | time$^{-1}$ |
| $a$ | Parameter | Conversion rate of carbon given to crab to crab defensiveness | time$^{-1}$ |
| $s$ | Parameter | Scaling factor for defensiveness capacity relative to $H$ | dimensionless |
| $\omega$ | Parameter | Defensiveness decay rate | time$^{-1}$ |

- Why have a weakening term? I think so that defensiveness can decrease if C gets too low--i.e., represents some cost to maintaining defensiveness; means D has an equilibrium where C-fueled growth balances general decay

# Simulate the model through time

Parameters

```{r}
r <- 0.3   # coral growth from carbon
l <- 0.2   # coral mortality rate ("loss")
p <- 0.4   # predation probability
e <- 0.5   # Endosymbiont carbon translocation rate 
h <- 0.2   # heterotrophic carbon acquisition rate
m <- 0.1  # maintenance carbon cost per unit coral
f <- 0.3   # coral carbon allocation to crabs
a <- 0.1   # conversion of carbon food to defensive behavior by crabs
s <- 2.0   # dcaling factor for crab defensiveness capacity relative to H
w <- 0.15  # defensiveness weakening/decay rate
```

Create holding vectors

```{r}
# create a vector of timesteps to iterate over
tset <- seq(from = 0, to = 10000, length.out = 5000)

# create holding vectors and store initial conditions (arbitrarily chosen)
H.simu <- NaN*tset; H.simu[1] <- 100
C.simu <- NaN*tset; C.simu[1] <- 100
D.simu <- NaN*tset; D.simu[1] <- 10
```

Run the simulation loop

```{r}
# for each timestep
for(i in 2:length(tset)){
  # calculate change in time
	dt <- tset[i] - tset[i-1]
	
	# store placeholder variables
	H <- H.simu[i-1]
	C <- C.simu[i-1]
	D <- D.simu[i-1]
	
	# calculate change in state variables
	dH <- ( r*C - l*H - p*H*(1/(1 + D)) )*dt
	dC <- ( e*H + h*H - r*C - m*H - f*C )*dt
	dD <- ( a*f*C * ((s*H - D)/(s*H)) - w*D )*dt
	
	# calculate total population size
	H.simu[i] <- H + dH
	C.simu[i] <- C + dC	
	D.simu[i] <- D + dD	
}
```

# Plot timeseries

```{r}
plot(x = tset, y = H.simu, 
     type = 'l', lwd = 2, lty = 1, col = "coral",
     xlab = 'Time', ylab = 'Quantity')
lines(x = tset, y = C.simu, 
     type = 'l', lwd = 2, lty = 2, col = "red")
lines(x = tset, y = D.simu, 
     type = 'l', lwd = 2, lty = 3, col = "black")

legend("topleft",
       col = c("coral", "red", "black"),
       lty = c(1, 2, 3),
       legend = c("Coral", "Carbon", "Defense"))
```

Seems like we need a logistic modifier?

# Logistic model

Now:

$$\frac{dH}{dt} = rC \left(1 - \frac{H}{K}\right) - pH\left(\frac{1}{1+D}\right)$$
Parameters (now no l, but have a K)

```{r}
r <- 0.3   # coral growth from carbon
K <- 200   # coral carrying capacity
p <- 0.6   # predation probability
e <- 0.5   # Endosymbiont carbon translocation rate 
h <- 0.2   # heterotrophic carbon acquisition rate
m <- 0.1  # maintenance carbon cost per unit coral
f <- 0.4   # coral carbon allocation to crabs
a <- 0.4   # conversion of carbon food to defensive behavior by crabs
s <- 2.0   # scaling factor for crab defensiveness capacity relative to H
w <- 0.15  # defensiveness weakening/decay rate
```

Create holding vectors

```{r}
# create a vector of timesteps to iterate over
tset <- seq(from = 0, to = 100, length.out = 5000)

# create holding vectors and store initial conditions (arbitrarily chosen)
H.simu <- NaN*tset; H.simu[1] <- 100
C.simu <- NaN*tset; C.simu[1] <- 100
D.simu <- NaN*tset; D.simu[1] <- 10
```

Run the simulation loop

```{r}
# for each timestep
for(i in 2:length(tset)){
  # calculate change in time
	dt <- tset[i] - tset[i-1]
	
	# store placeholder variables
	H <- H.simu[i-1]
	C <- C.simu[i-1]
	D <- D.simu[i-1]
	
	# calculate change in state variables
	dH <- ( r*C*(1 - (H/K)) - p*H*(1/(1 + D)) )*dt
	dC <- ( e*H + h*H - r*C - m*H - f*C )*dt
	dD <- ( a*f*C * ((s*H - D)/(s*H)) - w*D )*dt
	
	# calculate total population size
	H.simu[i] <- H + dH
	C.simu[i] <- C + dC	
	D.simu[i] <- D + dD	
}
```

# Plot timeseries

```{r}
plot(x = tset, y = H.simu, 
     type = 'l', lwd = 2, lty = 1, col = "coral",
     xlab = 'Time', ylab = 'Quantity',
     ylim = c(0, K))
lines(x = tset, y = C.simu, 
     type = 'l', lwd = 2, lty = 2, col = "red")
lines(x = tset, y = D.simu, 
     type = 'l', lwd = 2, lty = 3, col = "black")

legend("bottomright",
       col = c("coral", "red", "black"),
       lty = c(1, 2, 3),
       legend = c("Coral", "Carbon", "Defense"))
```

# Bifurcation of f vs. H*

```{r}
# values of f to iterate over
fset <- seq(from = 0, to = 2, length.out = 50) 

# holding vectors for equilibria
Hstarset <- NaN*fset
Cstarset <- NaN*fset
Dstarset <- NaN*fset

for(j in 1:length(fset)){
  f <- fset[j]
  
  H.simu <- NaN*tset; H.simu[1] <- 100
  C.simu <- NaN*tset; C.simu[1] <- 100
  D.simu <- NaN*tset; D.simu[1] <- 50
  
  
  for(i in 2:length(tset)){
    dt <- tset[i]-tset[i-1]
    # store placeholder variables
  	H <- H.simu[i-1]
  	C <- C.simu[i-1]
  	D <- D.simu[i-1]
  	
  	# calculate change in state variables
  	dH <- ( r*C*(1 - (H/K)) - p*H*(1/(1 + D)) )*dt
  	dC <- ( e*H + h*H - r*C - m*H - f*C )*dt
  	dD <- ( a*f*C * ((s*H - D)/(s*H)) - w*D )*dt
  	
  	# calculate total population size
  	H.simu[i] <- H + dH
  	C.simu[i] <- C + dC	
  	D.simu[i] <- D + dD	
  }
  
  Hstarset[j] <- H.simu[length(tset)]
  Cstarset[j] <- C.simu[length(tset)]
  Dstarset[j] <- D.simu[length(tset)]
}
```

Plot

```{r}
plot(x = fset, y = Hstarset, 
     type = 'l', lwd = 2, lty = 1, col = "coral",
     xlab = 'Allocation rate to crabs', ylab = 'Quantity',
     ylim = c(0, 350))
lines(x = fset, y = Cstarset, 
     type = 'l', lwd = 2, lty = 2, col = "red")
lines(x = fset, y = Dstarset, 
     type = 'l', lwd = 2, lty = 3, col = "black")

legend("topright",
       col = c("coral", "red", "black"),
       lty = c(1, 2, 3),
       legend = c("Coral", "Carbon", "Defense"))
```


